A Formal Model of Checked C

### 2023-01-13 sandp23/formal.tex
update.

\myparagraph{Types}
At a high level, we classify types as word-size value, multi-word
value, or function types. A word-size value can be either integer or pointer.
Every pointer type ($\tptr{\omega}{\xi}$) includes a
pointer mode annotation ($\xi$, the difference between context and pointer modes
is introduced shortly below) that is either checked (\cmode), tainted (\tmode),
or unchecked (\umode), and a type ($\omega$) denoting the valid value type it is pointed to.
A multi-word value type ($ \tallarrayb{\bvar}{\tau}$) that ranges over 
arrays and null-terminated arrays is constructed by the type of elements in the
array ($\tau$), an array bound ($\bvar$) comprised of an upper and
lower bound on the size of the array ($(b_l,b_h)$), and an array flag ($\kappa$).
Bounds $b$ are
limited to integer literals $n$ and expressions $x + n$.
Whether an array pointer is null terminated or not is determined by annotation
$\kappa$, which is $nt$ for null-terminated arrays, and $\cdot$
otherwise (we elide $\cdot$ when writing types).
An example representation of an array and null-terminated array in~\lang is shown below:
% 
\[\hspace*{-0.5em}
\begin{array}{l}
\begin{array}{rcl}
$\code{_t_Array_ptr<}$\tau$\code{> : count(}$n$\code{)}$
&\Leftrightarrow& \tarrayptr{0}{n}{\tau}{\tmode}
\\[0.2em]
$\code{_NT_Array_ptr<}$\tau$\code{> : count(}$n$\code{)}$
&\Leftrightarrow& \tntarrayptr{0}{n}{\tau}{\cmode}
\end{array}
\end{array}
\]

@@

%The \lang syntax of is given in Fig.~\ref{fig:checkc-syn}.
%There are two type notions in \lang.  Types $\tau$ classify
%word-sized values including integers and pointers, while types
%$\omega$ classify multi-word values such as arrays, null-terminated
%arrays, functions, and single-word-size values. 
%Pointer types ($\tptr{\omega}{\xi}$) include a pointer mode annotation 
%($\xi$, the difference between context and pointer modes is introduced shortly below)
%that is either checked (\cmode), tainted (\tmode), or unchecked (\umode),
%and a type ($\omega$) denoting valid values that can be pointed to.
%Array types include both the type of
%elements ($\tau$) and a bound ($\bvar$) comprised of an upper and
%lower bound on the size of the array ($(b_l,b_h)$). Bounds $b$ are
%limited to integer literals $n$ and expressions $x + n$.
%Whether an array pointer is null terminated or not is determined by annotation
%$\kappa$, which is $nt$ for null-terminated arrays, and $\cdot$
%otherwise (we elide $\cdot$ when writing types).
%
%\lang{} function types ($\tfun{\overline{x}}{\overline{\tau}}{\tau}$)
%reflect its dependent function declarations,
%where $\overline{x}$ represents
%a list of \tint{} type variables in a dependent function header
%that bind bound variables appearing in $\overline{\tau}$ and $\tau$.
%We have a well-formed requirement for a function type;
%that is, all variables in $\overline{\tau}$ and $\tau$ are bounded by $\overline{x}$.
%Here is the
%corresponding \systemname syntax for these types:
%\[\hspace*{-0.5em}
%\begin{array}{l}
%\begin{array}{rcl}
%$\code{_t_Array_ptr<}$\tau$\code{> : count(}$n$\code{)}$
%&\Leftrightarrow& \tarrayptr{0}{n}{\tau}{\tmode}
%\\[0.2em]
%$\code{_NT_Array_ptr<}$\tau$\code{> : count(}$n$\code{)}$
%&\Leftrightarrow& \tntarrayptr{0}{n}{\tau}{\cmode}
%\end{array}
%\\[0.2em]
%$\code{_t_ptr<(int)(_t_NT_Array_ptr<}$\tau$\code{> : count(}$n$\code{),}$\\
%\qquad\qquad$\code{_NT_Array_ptr<}$\tau$\code{>)>: count(}$n$\code{))>}$
%\\[0.2em]
%\Leftrightarrow\;\; $\tptr{(\tfun{n}{ \tntarrayptr{0}{n}{\tau}{\tmode} \times \tntarrayptr{0}{n}{\tau}{\tmode}}{\tint})}{\tmode}$
%\end{array}
%\]
%As a convention we write $\tptr{\tarrayb{b}{\tau}}{\cmode}$ to mean
%$\tptr{\tarray{0}{b}{\tau}}{\cmode}$, so the firs two examples above could
%be rewritten $\tptr{\tarrayb{n}{\tau}}{\cmode}$ and
%$\tptr{\tntarrayb{n}{\tau}}{\cmode}$, respectively.
\myparagraph{Expressions}
\lang expressions include common expressions such as addition ($\ebinop{e_1}{e_2}$)
\mzu{($n\!:\!\tau$)}\mz{what's this?}, pointer dereference and assignment ($\estar{e}$) and ($\eassign{e_1}{e_2}$), resp.
along with expressions that require special handling, such as,
 static casts ($\ecast{\tau}{e}$), dynamic casts ($\edyncast{\tau}{e}$) \footnote{assumed at compile-time and verified at run-time, see \Cref{app:main}}, the \texttt{strlen} operation ($\estrlen{x}$),
memory allocation ($\emalloc{\xi}{\omega}$), 
function calls ($\ecall{e}{\overline{e}}$),
unchecked blocks ($\eunchecked{\overline{x}}{e}$), and checked blocks ($\echecked{\overline{x}}{e}$).
For instance, a dynamic bounds cast expression
\begin{center}
  {\footnotesize
    \code{dyn_bounds_cast<_Array_ptr<}$\tau$\code{>>(}$e$\code{,
      count(}$n$\code{))} }
\end{center}
becomes 
{\footnotesize$\edyncast{\tptr{\tarrayb{n}{\tau}}{\cmode}}{e}$}. 
%

% 
\mzu{We denote integer literals $n$ with a type $\tau$ (\ie $\tint$ or $\tptr{\omega}{\xi}$), enabling the use of fixed addresses as pointers.}
\mz{This sentence is unclear. I suppose that you mean an integer literal
  can be both of interger type and pointer type depending on the context.  }
For instance, $\evalue{0}{\tptr{\omega}{\xi}}$ (for any $\xi$ and $\omega$) represents the $\enull$ pointer.
The~\code{strlen} expression plays an important role in the type-checking of null-terminated arrays.
For simplifying type checking, we assume~\code{strlen} to operate on variables \mzs{$x$} rather than arbitrary expressions.
% 
The heap allocation expression $\emalloc{\xi}{\omega}$ includes a mode flag $\xi$ for allocating memory in different regions (\cregion or \ucregion).
We disallow $\omega$ to be a function type ($\tfun{\overline{x}}{\overline{\tau}}{\tau}$).
The $\echeckedtext$ and $\euncheckedtext$ expressions are used to
\mzs{represent the corresponding} \mzr{delimit} code regions.
To guarantee the non-exposure property, we extend the base syntax to be $\echecked{\overline{x}}{e}$ and $\eunchecked{\overline{x}}{e}$.
\mzu{Wherein $\overline{x}$ restricts all free variables appearing in $e$ to be checked or unchecked pointers according to the corresponding region.}
\mz{Is this sentence grammatical?}
%

% 
\lang{} aims to be simple enough to work with but powerful enough to
encode realistic \systemname idioms. For example, mutable local
  variables can be encoded as immutable locals that point to the heap;
\mzu{the use of \code{&} can be simulated with \code{malloc};}
\mz{This doesn't make sense if you use \code{&} to create alias. You don't want
  to allocate memory.}
% 
and loops can be encoded as recursive function calls. \code{struct}s are
not included in~\fig{fig:checkc-syn} for space reasons, but they are
supported as shown in~\apdx{appx:struct}.
C-style \code{union}s have no safe typing in \checkedc, so we omit them.

Although the base syntax of~\lang is similar to that of~\checkedc model~\cite{li22checkedc}, there are considerable enhancements to support tainted types ($t$ in $\xi$), special heap handling (\ie $\emalloc{\xi}{\omega}$), and explicit specification of pointers in the $\echeckedtext$ and $\euncheckedtext$ regions.

@@

\aravind{Why we have only function pointers in function call?}
\aravind{Talk about the ret expression after adding it in the syntax.}

@@

\begin{figure}
{\small
$\hspace*{-1.2em}
    \begin{array}{l}
    \begin{array}{lll}
e & ::= & \ldots \mid \ret{x}{\evalue{n}{\tau}}{e}\\
r & ::= & e \mid \enull \mid \ebounds\\
E &::=& \Box \mid \ebinop{E}{e} \mid \ebinop{\evalue{n}{\tau}}{E}\mid \ecast{\tau}{E} \mid \edyncast{\tau}{E} \mid\estar{E}\mid\eassign{E}{e}\\[0.2em]
&&\mid\eassign{\evalue{n}{\tau}}{E}\mid \elet{x}{E}{e}\mid \eif{E}{e}{e}\\[0.2em]
&&\mid \ecall{E}{\overline{e}} \mid \ecall{\evalue{n}{\tau}}{\overline{E}} \mid 
\eunchecked{\overline{x}}{E}
\mid \echecked{\overline{x}}{E}

>>>>

\myparagraph{Type Syntax}
At a high level, we classify types as word-size value, multi-word
value, or function types. A word-size value can be either integer or pointer.
Every pointer type ($\tptr{\omega}{\xi}$) includes a
pointer mode annotation ($\xi$, the difference between context and pointer modes
is introduced shortly below) that is either checked (\cmode), tainted (\tmode),
or unchecked (\umode), and a type ($\omega$) denoting the valid value type it is pointed to.
A multi-word value type ($ \tallarrayb{\bvar}{\tau}$) that ranges over 
arrays and null-terminated arrays is constructed by the type of elements in the
array ($\tau$), an array bound ($\bvar$) comprised of an upper and
lower bound on the size of the array ($(b_l,b_h)$), and an array flag ($\kappa$).
Bounds $b$ are
limited to integer literals $n$ and expressions $x + n$.
Whether an array pointer is null terminated or not is determined by annotation
$\kappa$, which is $nt$ for null-terminated arrays, and $\cdot$
otherwise (we elide $\cdot$ when writing types).
An example representation of an array and null-terminated array in~\lang is shown below:
% 
\[\hspace*{-0.5em}
\begin{array}{l}
\begin{array}{rcl}
$\code{_t_Array_ptr<}$\tau$\code{> : count(}$n$\code{)}$
&\Leftrightarrow& \tarrayptr{0}{n}{\tau}{\tmode}
\\[0.2em]
$\code{_NT_Array_ptr<}$\tau$\code{> : count(}$n$\code{)}$
&\Leftrightarrow& \tntarrayptr{0}{n}{\tau}{\cmode}
\end{array}
\end{array}
\]

@@

%The \lang syntax of is given in Fig.~\ref{fig:checkc-syn}.
%There are two type notions in \lang.  Types $\tau$ classify
%word-sized values including integers and pointers, while types
%$\omega$ classify multi-word values such as arrays, null-terminated
%arrays, functions, and single-word-size values. 
%Pointer types ($\tptr{\omega}{\xi}$) include a pointer mode annotation 
%($\xi$, the difference between context and pointer modes is introduced shortly below)
%that is either checked (\cmode), tainted (\tmode), or unchecked (\umode),
%and a type ($\omega$) denoting valid values that can be pointed to.
%Array types include both the type of
%elements ($\tau$) and a bound ($\bvar$) comprised of an upper and
%lower bound on the size of the array ($(b_l,b_h)$). Bounds $b$ are
%limited to integer literals $n$ and expressions $x + n$.
%Whether an array pointer is null terminated or not is determined by annotation
%$\kappa$, which is $nt$ for null-terminated arrays, and $\cdot$
%otherwise (we elide $\cdot$ when writing types).
%
%\lang{} function types ($\tfun{\overline{x}}{\overline{\tau}}{\tau}$)
%reflect its dependent function declarations,
%where $\overline{x}$ represents
%a list of \tint{} type variables in a dependent function header
%that bind bound variables appearing in $\overline{\tau}$ and $\tau$.
%We have a well-formed requirement for a function type;
%that is, all variables in $\overline{\tau}$ and $\tau$ are bounded by $\overline{x}$.
%Here is the
%corresponding \systemname syntax for these types:
%\[\hspace*{-0.5em}
%\begin{array}{l}
%\begin{array}{rcl}
%$\code{_t_Array_ptr<}$\tau$\code{> : count(}$n$\code{)}$
%&\Leftrightarrow& \tarrayptr{0}{n}{\tau}{\tmode}
%\\[0.2em]
%$\code{_NT_Array_ptr<}$\tau$\code{> : count(}$n$\code{)}$
%&\Leftrightarrow& \tntarrayptr{0}{n}{\tau}{\cmode}
%\end{array}
%\\[0.2em]
%$\code{_t_ptr<(int)(_t_NT_Array_ptr<}$\tau$\code{> : count(}$n$\code{),}$\\
%\qquad\qquad$\code{_NT_Array_ptr<}$\tau$\code{>)>: count(}$n$\code{))>}$
%\\[0.2em]
%\Leftrightarrow\;\; $\tptr{(\tfun{n}{ \tntarrayptr{0}{n}{\tau}{\tmode} \times \tntarrayptr{0}{n}{\tau}{\tmode}}{\tint})}{\tmode}$
%\end{array}
%\]
%As a convention we write $\tptr{\tarrayb{b}{\tau}}{\cmode}$ to mean
%$\tptr{\tarray{0}{b}{\tau}}{\cmode}$, so the firs two examples above could
%be rewritten $\tptr{\tarrayb{n}{\tau}}{\cmode}$ and
%$\tptr{\tntarrayb{n}{\tau}}{\cmode}$, respectively.
\myparagraph{Expressions}
\lang expressions include common expressions such as addition ($\ebinop{e_1}{e_2}$), 
pointer dereference ($\estar{e}$) and assignment ($\eassign{e_1}{e_2}$),
along with expressions that require special handling, such as,
 static casts ($\ecast{\tau}{e}$), dynamic casts ($\edyncast{\tau}{e}$) \footnote{assumed at compile-time and verified at run-time, see \Cref{app:main}}, the \texttt{strlen} operation ($\estrlen{x}$),
memory allocation ($\emalloc{\xi}{\omega}$), 
function calls ($\ecall{e}{\overline{e}}$),
unchecked blocks ($\eunchecked{\overline{x}}{e}$), and checked blocks ($\echecked{\overline{x}}{e}$).
For example, a dynamic bounds cast
  {\footnotesize
    \code{dyn_bounds_cast<_Array_ptr<}$\tau$\code{>>(}$e$\code{,count(}$n$\code{))} }
becomes 
{\footnotesize$\edyncast{\tptr{\tarrayb{n}{\tau}}{\cmode}}{e}$}. 
%

% 
We denote integer literals $n$ with a type $\tau$ (\ie $\tint$ or $\tptr{\omega}{\xi}$), enabling the use of fixed addresses as pointers.
For instance, $\evalue{0}{\tptr{\omega}{\xi}}$ (for any $\xi$ and $\omega$) represents the $\enull$ pointer.
The heap allocation expression $\emalloc{\xi}{\omega}$ includes a mode flag $\xi$ for allocating memory in different regions (\cregion or \ucregion).
We disallow $\omega$ to be a function type ($\tfun{\overline{x}}{\overline{\tau}}{\tau}$).
The $\echeckedtext$ and $\euncheckedtext$ expressions are used to delimit code regions.
To guarantee the non-exposure property, we extend the \checkedc syntax to be $\echecked{\overline{x}}{e}$ and $\eunchecked{\overline{x}}{e}$,
where $\overline{x}$ represents all variables that are allowed to communicate between instructions outside and inside of the block $e$, and cannot contain checked pointers.
$\erettext$ is introduced by the semantics when evaluating a \texttt{let} binding; explained shortly below.

% 
\lang{} aims to be simple enough to work with but powerful enough to
encode realistic \systemname idioms. For example,
loops can be encoded as recursive function calls. \code{struct}s are
not included in~\fig{fig:checkc-syn} for space reasons, but they are
supported as shown in~\cite{li22checkedc}.
C-style \code{union}s have no safe typing in \checkedc, so we omit them.
Although the base syntax of~\lang is similar to that of~\checkedc model~\cite{li22checkedc}, there are considerable enhancements to support tainted types ($t$ in $\xi$), special heap handling (\ie $\emalloc{\xi}{\omega}$), and explicit specification of pointers in the $\echeckedtext$ and $\euncheckedtext$ regions.

@@

@@

\begin{figure}
{\small
$\hspace*{-1.2em}
    \begin{array}{l}
    \begin{array}{lll}
r & ::= & e \mid \enull \mid \ebounds\\
E &::=& \Box \mid \ebinop{E}{e} \mid \ebinop{\evalue{n}{\tau}}{E}\mid \ecast{\tau}{E} \mid \edyncast{\tau}{E} \mid\estar{E}\mid\eassign{E}{e}\\[0.2em]
&&\mid\eassign{\evalue{n}{\tau}}{E}\mid \elet{x}{E}{e}\mid \eif{E}{e}{e}\\[0.2em]
&&\mid \ecall{E}{\overline{e}} \mid \ecall{\evalue{n}{\tau}}{\overline{E}} \mid 
\eunchecked{\overline{x}}{E}
\mid \echecked{\overline{x}}{E}
